### 2023-09-01 proofs/tex/subst-lemmas.tex
Finish syntax update

\begin{lemma}[Algorithmic Substitution strengthening]
  \label{lemma:subst-restr-uv}
  Restricting the substitution to the free algorithmic 
  variables of the
  substitution subject does not affect the result.
  Suppose that $[[uσ]]$ is an algorithmic substitution,
  $[[uP]]$ and $[[uN]]$ are algorithmic types. Then
 \begin{itemize}
    \item[$+$] $[[ [uσ]uP ]] = [[ [uσ|uv uP]uP ]]$,
    \item[$-$] $[[ [uσ]uN ]] = [[ [uσ|uv iN]uN ]]$
  \end{itemize}
\end{lemma}
\begin{proof}
  The proof is analogous to the proof of \cref{lemma:subst-restr-fv}.
\end{proof}

@@

\begin{lemma}[Substitutions equal on the metavariables]
  \label{lemma:subst-eq-metavar}
  Suppose that $[[uσ1]]$ and $[[uσ2]]$ are 
  normalized substitutions of signature $[[Θ ⊢ uσi]]$.
  Then 
  \begin{enumerate}
    \item [$+$] for a normalized type $[[Γ; dom(Θ) ⊢ uP]]$, if $[[ [uσ1]uP = [uσ2]uP ]]$ then
      $[[uσ1|(uv uP)  = uσ2|(uv uP)]]$;
    \item [$-$] for a normalized type $[[Γ; dom(Θ) ⊢ uN]]$, if $[[ [uσ1]uN = [uσ2]uN ]]$ then
      $[[uσ1|(uv uN)  = uσ2|(uv uN)]]$.
  \end{enumerate}
\end{lemma}
\begin{proof}
  The proof is a simple structural induction on 
  $[[Γ; dom(Θ) ⊢ uP]]$ and mutually, on $[[Γ; dom(Θ) ⊢ uN]]$.
  Let us consider the shape of $[[uN]]$ (the cases of $[[uP]]$ are symmetric).
  \begin{caseof}
    \item $[[Γ; dom(Θ) ⊢ α̂⁻]]$. Then $[[ [uσ1]α̂⁻ = [uσ2]α̂⁻ ]]$
      implies $[[uσ1|(uv α̂⁻) = uσ2|(uv α̂⁻)]]$ immediately.
    \item $[[Γ; dom(Θ) ⊢ α⁻]]$. Then $[[uv α̂⁻ = ∅]]$, and
      $[[uσ1|(uv α̂⁻)  = uσ2|(uv α̂⁻)]]$ holds vacuously.
    \item $[[Γ; dom(Θ) ⊢ ∀pas.uN]]$. Then we are proving that
      $[[ [uσ1]∀pas.uN = [uσ2]∀pas.uN ]]$ implies $[[uσ1|(uv ∀pas.uN) = uσ2|(uv ∀pas.uN)]]$.
      By definition of substitution and \ruleref{\ottdruleEOneForallLabel}, 
      $[[ [uσ1]uN = [uσ2]uN ]]$ implies $[[uσ1|uv uN = uσ2|uv uN]]$.
      Since $[[∀pas.uN]]$ is normalized, so is $[[Γ, pas; dom(Θ) ⊢ uN]]$, 
      hence, the induction hypothesis is applicable and implies $[[uσ1|uv uN = uσ2|uv uN]]$,
      as required.
    \item $[[Γ; dom(Θ) ⊢ uP → uN]]$. Then we are proving that
      $[[ [uσ1](uP → uN) = [uσ2](uP → uN) ]]$ implies $[[uσ1|(uv uP → uN) = uσ2|(uv uP → uN)]]$.
      By definition of substitution and congruence of equality, 
      $[[ [uσ1](uP → uN) = [uσ2](uP → uN) ]]$
      means $[[ [uσ1]uP = [uσ2]uP ]]$ and $[[ [uσ1]uN = [uσ2]uN ]]$.
      Notice that $[[uP]]$ and $[[uN]]$ are normalized since $[[uP → uN]]$ is normalized, 
      and well-formed in the same contexts.
      This way, by the induction hypothesis, 
      $[[uσ1|(uv uP) = uσ2|(uv uP)]]$ and $[[uσ1|(uv uN) = uσ2|(uv uN)]]$,
      which since $[[uv (uP → uN) = uv uP ∪ uv uN]]$ implies
      $[[uσ1|(uv uP → uN) = uσ2|(uv uP → uN)]]$.
    \item $[[Γ; dom(Θ) ⊢ ↑uP]]$. The proof is similar to the previous case:
      we apply congruence of substitution, equality, and normalization,
      then the induction hypothesis, and then the fact that $[[uv (↑uP) = uv uP]]$.
  \end{caseof}
\end{proof}

\begin{corollary}[Substitutions equivalent on the metavariables]
  \label{lemma:subst-equiv-metavar}
  Suppose that $[[uσ1]]$ and $[[uσ2]]$ are substitutions 
  of signature $[[Θ ⊢ uσi]]$.
  Then 
  \begin{enumerate}
    \item [$+$] for a type $[[Γ; dom(Θ) ⊢ uP]]$, if $[[Θ ⊢ [uσ1]uP ≈ [uσ2]uP]]$ then
      $[[Θ|uv uP ⊢ uσ1 ≈ uσ2]]$;
    \item [$-$] for a type $[[Γ; dom(Θ) ⊢ uN]]$, if $[[Θ ⊢ [uσ1]uN ≈ [uσ2]uN]]$ then
      $[[Θ|uv uN ⊢ uσ1 ≈ uσ2]]$.
  \end{enumerate}
\end{corollary}
\begin{proof}
  First, let us normalize the types and the substitutions, and show that
  the given equivalences and well-formedness properties are preserved. 
  $[[Γ; dom(Θ) ⊢ uP]]$ implies $[[Γ; dom(Θ) ⊢ nf(uP)]]$ by \cref{corollary:wf-nf-algo}.
  $[[Θ ⊢ [uσ1]uP ≈ [uσ2]uP]]$ implies $[[ nf([uσ1]uP) = nf([uσ2]uP)]]$ by 
  \cref{lemma:subt-equiv-algorithmization}.
  Then $[[ nf([uσ1]uP) = nf([uσ2]uP)]]$ implies 
  $[[ [nf(uσ1)]nf(uP) = [nf(uσ2)]nf(uP)]]$ by \cref{lemma:norm-subst-distr}.
  Notice that by \cref{corollary:norm-subst-sig-algo}
  $[[Θ ⊢ uσi]]$ implies $[[Θ ⊢ nf(uσi)]]$.

  This way, by \cref{lemma:subst-eq-metavar}, 
  $[[Θ ⊢ [uσ1]uP ≈ [uσ2]uP]]$ implies 
  $[[nf(uσ1)|(uv nf(uP)) = nf(uσ2)|(uv nf(uP))]]$.
  Then by \cref{lemma:uv-nf}, 
  $[[nf(uσ1)|(uv uP) = nf(uσ2)|(uv uP)]]$,
  and by ${{corollary:subst-subt-equiv-algorithmization}}$,
  $[[Θ|uv uP ⊢ uσ1 ≈ uσ2]]$.

  Symmetrically, 
  $[[Θ ⊢ [uσ1]uN ≈ [uσ2]uN]]$ implies 
  $[[Θ|uv uN ⊢ uσ1 ≈ uσ2]]$.
\end{proof}

>>>>

\begin{lemma}[Algorithmic Substitution Strengthening]
  \label{lemma:subst-restr-uv}
  Restricting the substitution to the free algorithmic 
  variables of the
  substitution subject does not affect the result.
  Suppose that $[[uσ]]$ is an algorithmic substitution,
  $[[uP]]$ and $[[uN]]$ are algorithmic types. Then
 \begin{itemize}
    \item[$+$] $[[ [uσ]uP ]] = [[ [uσ|uv uP]uP ]]$,
    \item[$-$] $[[ [uσ]uN ]] = [[ [uσ|uv uN]uN ]]$
  \end{itemize}
\end{lemma}
\begin{proof}
  The proof is analogous to the proof of \cref{lemma:subst-restr-fv}.
\end{proof}

@@

\begin{lemma}[Substitutions equal on the metavariables]
  \label{lemma:subst-eq-metavar}
  Suppose that $[[uσ1]]$ and $[[uσ2]]$ are 
  normalized substitutions of signature $[[Θ ⊢ uσi : Ξ]]$.
  Then 
  \begin{enumerate}
    \item [$+$] for a normalized type $[[Γ; Ξ ⊢ uP]]$, if $[[ [uσ1]uP = [uσ2]uP ]]$ then
      $[[uσ1|(uv uP)  = uσ2|(uv uP)]]$;
    \item [$-$] for a normalized type $[[Γ; Ξ ⊢ uN]]$, if $[[ [uσ1]uN = [uσ2]uN ]]$ then
      $[[uσ1|(uv uN)  = uσ2|(uv uN)]]$.
  \end{enumerate}
\end{lemma}
\begin{proof}
  The proof is a simple structural induction on 
  $[[Γ; Ξ ⊢ uP]]$ and mutually, on $[[Γ; Ξ ⊢ uN]]$.
  Let us consider the shape of $[[uN]]$ (the cases of $[[uP]]$ are symmetric).
  \begin{caseof}
    \item $[[Γ; Ξ ⊢ α̂⁻]]$. Then $[[ [uσ1]α̂⁻ = [uσ2]α̂⁻ ]]$
      implies $[[uσ1|(uv α̂⁻) = uσ2|(uv α̂⁻)]]$ immediately.
    \item $[[Γ; Ξ ⊢ α⁻]]$. Then $[[uv α̂⁻ = ∅]]$, and
      $[[uσ1|(uv α̂⁻)  = uσ2|(uv α̂⁻)]]$ holds vacuously.
    \item $[[Γ; Ξ ⊢ ∀pas.uN]]$. Then we are proving that
      $[[ [uσ1]∀pas.uN = [uσ2]∀pas.uN ]]$ implies $[[uσ1|(uv ∀pas.uN) = uσ2|(uv ∀pas.uN)]]$.
      By definition of substitution and \ruleref{\ottdruleEOneForallLabel}, 
      $[[ [uσ1]uN = [uσ2]uN ]]$ implies $[[uσ1|uv uN = uσ2|uv uN]]$.
      Since $[[∀pas.uN]]$ is normalized, so is $[[Γ, pas; Ξ ⊢ uN]]$, 
      hence, the induction hypothesis is applicable and implies $[[uσ1|uv uN = uσ2|uv uN]]$,
      as required.
    \item $[[Γ; Ξ ⊢ uP → uN]]$. Then we are proving that
      $[[ [uσ1](uP → uN) = [uσ2](uP → uN) ]]$ implies $[[uσ1|(uv uP → uN) = uσ2|(uv uP → uN)]]$.
      By definition of substitution and congruence of equality, 
      $[[ [uσ1](uP → uN) = [uσ2](uP → uN) ]]$
      means $[[ [uσ1]uP = [uσ2]uP ]]$ and $[[ [uσ1]uN = [uσ2]uN ]]$.
      Notice that $[[uP]]$ and $[[uN]]$ are normalized since $[[uP → uN]]$ is normalized, 
      and well-formed in the same contexts.
      This way, by the induction hypothesis, 
      $[[uσ1|(uv uP) = uσ2|(uv uP)]]$ and $[[uσ1|(uv uN) = uσ2|(uv uN)]]$,
      which since $[[uv (uP → uN) = uv uP ∪ uv uN]]$ implies
      $[[uσ1|(uv uP → uN) = uσ2|(uv uP → uN)]]$.
    \item $[[Γ; Ξ ⊢ ↑uP]]$. The proof is similar to the previous case:
      we apply congruence of substitution, equality, and normalization,
      then the induction hypothesis, and then the fact that $[[uv (↑uP) = uv uP]]$.
  \end{caseof}
\end{proof}

\begin{corollary}[Substitutions equivalent on the metavariables]
  \label{lemma:subst-equiv-metavar}
  Suppose that $[[uσ1]]$ and $[[uσ2]]$ are substitutions 
  of signature $[[Θ ⊢ uσi : Ξ]]$.
  Then 
  \begin{enumerate}
    \item [$+$] for a type $[[Γ; Ξ ⊢ uP]]$, if $[[Θ ⊢ [uσ1]uP ≈ [uσ2]uP]]$ then
      $[[Θ ⊢ uσ1 ≈ uσ2 : uv uP]]$;
    \item [$-$] for a type $[[Γ; Ξ ⊢ uN]]$, if $[[Θ ⊢ [uσ1]uN ≈ [uσ2]uN]]$ then
      $[[Θ ⊢ uσ1 ≈ uσ2 : uv uN]]$.
  \end{enumerate}
\end{corollary}
\begin{proof}
  First, let us normalize the types and the substitutions, and show that
  the given equivalences and well-formedness properties are preserved. 
  $[[Γ; Ξ ⊢ uP]]$ implies $[[Γ; Ξ ⊢ nf(uP)]]$ by \cref{corollary:wf-nf-algo}.
  $[[Θ ⊢ [uσ1]uP ≈ [uσ2]uP]]$ implies $[[ nf([uσ1]uP) = nf([uσ2]uP)]]$ by 
  \cref{lemma:subt-equiv-algorithmization}.
  Then $[[ nf([uσ1]uP) = nf([uσ2]uP)]]$ implies 
  $[[ [nf(uσ1)]nf(uP) = [nf(uσ2)]nf(uP)]]$ by \cref{lemma:norm-subst-distr}.
  Notice that by \cref{corollary:norm-subst-sig-algo}
  $[[Θ ⊢ uσi : Ξ]]$ implies $[[Θ ⊢ nf(uσi) : Ξ]]$.

  This way, by \cref{lemma:subst-eq-metavar}, 
  $[[Θ ⊢ [uσ1]uP ≈ [uσ2]uP]]$ implies 
  $[[nf(uσ1)|(uv nf(uP)) = nf(uσ2)|(uv nf(uP))]]$.
  Then by \cref{lemma:uv-nf}, 
  $[[nf(uσ1)|(uv uP) = nf(uσ2)|(uv uP)]]$,
  and by ${{corollary:subst-subt-equiv-algorithmization}}$,
  $[[Θ ⊢ uσ1 ≈ uσ2 : uv uP]]$.

  Symmetrically, 
  $[[Θ ⊢ [uσ1]uN ≈ [uσ2]uN]]$ implies 
  $[[Θ ⊢ uσ1 ≈ uσ2 : uv uN]]$.
\end{proof}
